---
title: "ExploreData"
author: "John Phelan and Chris Madsen"
date: "`r Sys.Date()`"
output: html_document
params: 
  sppOI: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8)
library(tidyverse)
library(DBI)
library(bcinvadeR)
library(terra)
library(sf)
library(geodata)
library(predicts)
library(ggpubr)
library(dismo)
library(rJava)
library(spatstat)
library(ape)
library(bcmaps)
source("ZuurFuncs.R")
source("scripts/utils/thin_occ.R")

#set locations
proj_wd = getwd()
onedrive_wd = paste0(str_extract(getwd(),"C:/Users/[A-Z]+/"),"OneDrive - Government of BC/data/CNF/")

```

## Data description

The objectives of this work/study are to describe the highest risk locations for aquatic invasive species (AIS), by species and by identifying suitable habitat per species, a general map of potential suitable habitat for AIS. To do this, presence data is supplied to a model with a list of potential explanatory variables.

Presence data comes from a range of sources:

1.	Wildlife Species Inventory Incidental Observations layer on BC Warehouse

2.	Aquatic Invasive Species of British Columbia layer on BC Warehouse

3.	Master Incidence Report Records excel file

4.	iNaturalist (BC only)

These sources provide us with the source of the data, The date of recording, the species in question, the name of the location, and the geometry (location) where the species was recorded. There are no absence data supplied.

For explanatory variables, there are a number of data available. The bioclim model gives a range of climatic variable that can be included, such as temperature and precipitation. This is a raster file. The bioclim data is available both for present day and future climate scenarios. There are also a range of resolutions for the data. Resolutions may need to be altered for modelling. The data are supplied as rasters and are in 2.5, 5 or 10 minute increments. Water chemistry and temperature data are available from other sources, such as the environmental monitoring dataset.  

pH and Calcium data are available from Geurin et al. 2024 where they used water quality data to interpolate values for these variables for North America in a 10*10km grid. Data are supplied as rasters. This data are sourced in the EMS database too.

```{r data_collection, echo=FALSE}
# sppOI<-"goldfish"
sppOI = params$sppOI

sppOI<-"asian clam"

# Get occurrence data for species (if we haven't done this yet!)
if(!exists("d")){
  d = suppressMessages(grab_aq_occ_data(sppOI))
}

d_thin<-thin_occ(d)
# Pull in climate variables
ph_NAM = terra::rast(paste0(onedrive_wd,"ph-KR-208784-median_10km_ZN.tif")) |> terra::project("EPSG:4326")
names(ph_NAM) <- "pH"
Calc_NAM = terra::rast(paste0(onedrive_wd,"calcium-KR-97648-median-10km-ZN.tif")) |> terra::project("EPSG:4326")
names(Calc_NAM) <- "calc"

# Read in variable names for the worldclim variables.
source(paste0(proj_wd,"/scripts/utils/worldclim_bioc_var_names.R"))
cmidata<-cmip6_world("ACCESS-CM2", ssp = "585", var = "bioc", res = 5, time = "2021-2040",path="CMI/")
names(cmidata)<-renames
pred_bioc = terra::rast("CMI/climate/wc2.1_5m/wc2.1_5m_bioc_ACCESS-CM2_ssp585_2021-2040.tif")
names(pred_bioc)<-renames

# Cut our rasters down to just BC.
bc_vect = terra::vect(sf::st_transform(bcmaps::bc_bound(),4326))
ph_clipped = mask(crop(ph_NAM, bc_vect), bc_vect)
calc_clipped = mask(crop(Calc_NAM, bc_vect), bc_vect)
pred_bioc_clipped = mask(crop(pred_bioc, bc_vect), bc_vect)

# Pull in elevation raster with {elevatr}
elev = suppressMessages(elevatr::get_elev_raster(locations = sf::st_as_sf(bc_vect), z = 4)) |>
  terra::rast() |> 
  terra::crop(bc_vect) |> 
  terra::mask(bc_vect)
names(elev) = "elev"

```




```{r fixing_data}
NA_ph_res = terra::resample(ph_clipped, pred_bioc_clipped)
NA_calc_res = terra::resample(calc_clipped, pred_bioc_clipped)

elev_res = terra::resample(elev, pred_bioc_clipped)

all_rasters = c(pred_bioc_clipped, NA_ph_res, NA_calc_res, elev_res)

names(all_rasters)[c(1,12)] <- c("temp","precip")
temps<-pred_bioc_clipped[[1]]
precip<-pred_bioc_clipped[[12]]

stack_data<-c(NA_ph_res, NA_calc_res, elev_res, temps, precip)
names(stack_data)<-c("pH", "calc", "elev", "temp", "precip")

# Trim away bits outside of watercourses.
watercourses = terra::rast(paste0(onedrive_wd,"stream_density_6_plus_order_raster.tif")) |> 
  terra::project("EPSG:4326") |> 
  terra::resample(pred_bioc_clipped) |> 
  terra::mask(bc_vect)


samp<- d %>% 
    plyr::mutate(x = st_coordinates(geometry)[,1],
                     y = st_coordinates(geometry)[,2]) %>% 
    sf::st_drop_geometry() |> 
    dplyr::select(x,y) |> 
    mutate(present = 1)



samp_thin<-d_thin %>% 
  plyr::mutate(x = st_coordinates(geometry)[,1],
                     y = st_coordinates(geometry)[,2]) %>% 
    sf::st_drop_geometry() |> 
    dplyr::select(x,y) |> 
    mutate(present = 1)

```

```{r get_var_at_spp_occurance}
# Extract raster variable values at species occurrence point locations.
samp$temp = terra::extract(pred_bioc_clipped[[1]], samp[,c("x","y")], ID = FALSE)
samp$precip = terra::extract(pred_bioc_clipped$Annual_Precipitation, samp[,c("x","y")], ID = FALSE)
samp$calc = terra::extract(calc_clipped$calc, samp[,c("x","y")], ID = FALSE)
samp$pH = terra::extract(ph_clipped, samp[,c("x","y")], ID = FALSE)
samp$elev = terra::extract(elev, samp[,c("x","y")], ID = FALSE)
samp = as.data.frame(samp)

samp = samp |> 
  mutate(across(everything(), \(x) as.numeric(unlist(x))))
#dop NA's
samp = samp[complete.cases(samp),]


samp_thin$temp = terra::extract(pred_bioc_clipped[[1]], samp_thin[,c("x","y")], ID = FALSE)
samp_thin$precip = terra::extract(pred_bioc_clipped$Annual_Precipitation, samp_thin[,c("x","y")], ID = FALSE)
samp_thin$calc = terra::extract(calc_clipped$calc, samp_thin[,c("x","y")], ID = FALSE)
samp_thin$pH = terra::extract(ph_clipped, samp_thin[,c("x","y")], ID = FALSE)
samp_thin$elev = terra::extract(elev, samp_thin[,c("x","y")], ID = FALSE)
samp_thin = as.data.frame(samp_thin)

samp_thin = samp_thin |> 
  mutate(across(everything(), \(x) as.numeric(unlist(x))))
#dop NA's
samp_thin = samp_thin[complete.cases(samp_thin),]




```

### Where are the observations? {.tabset}

#### Presence plot
```{r plot_points}

bc<-sf::st_transform(bcmaps::bc_bound(),4326)

ggplot()+
  geom_sf(data = bc, color = "lightgrey", alpha = 0.7)+
  geom_point(data = samp, aes (x = x, y = y, color = as.factor(present)), size = 1.3)+
  scale_color_manual(values = c("0" = "purple", "1" = "darkgreen"),
                     labels = c("0" = "Absent", "1" = "Present"))+
  labs(color = paste0("Presence of \n",sppOI))+
  ggthemes::theme_map()+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = c(0.85, 0.5),
        panel.background = element_blank()
  ) 
  

```

#### Density plot
```{r occ_heatmap}
library(spatstat)

bc_alb<-st_transform(bc_bound(), 3005)

d_alb<-st_transform(d, 3005)

d_alb<- d_alb %>% 
    plyr::mutate(x = st_coordinates(geometry)[,1],
                     y = st_coordinates(geometry)[,2]) %>% 
    #sf::st_drop_geometry() |> 
    dplyr::select(x,y)


raster_base <- rast(vect(bc_alb))
res(raster_base)<-10000
values(raster_base)<-0
raster_occ <- rasterize(vect(d_alb), raster_base, fun = "count")
raster_occ_masked <- mask(raster_occ, raster_base)


kdePlot<-ggplot()+
  geom_sf(data= bc_alb)+
  stat_density2d(aes(x  = x, y = y, fill = after_stat(level)), alpha = .5,
                 h = 100000,n = 1000,geom = "polygon", data = d_alb)+ 
  scale_fill_viridis_c()+
  theme(legend.position = 'none')

ggsave("./images/kdePlot.png",kdePlot, height = 10, width = 12, units = "in")



bc_alb

library(spatstat)
ppp_d <- as.ppp(d_alb)

ppp_dens<-density(ppp_d, dimyx = 1000, bw = 1000 , kernel = "epanechnikov")
plot(ppp_dens)
plot(ppp_d, add = T)
Window(ppp_d) <- as.owin(st_transform(bc,3005))  
#plot(ppp_d, main = "")

density_stars <- stars::st_as_stars(ppp_dens)
density_sf <- st_as_sf(density_stars) %>%
  st_set_crs(3005)
  

ggplot() +
  geom_sf(data = density_sf, aes(fill = v), col = NA) +
  scale_fill_viridis_c() +
  geom_sf(data = st_boundary(bc_alb)) +
  geom_sf(data = st_transform(d,3005), size = 2, col = "black")

# x1<-terra::ext(bc_vect)[1];x2<-terra::ext(bc_vect)[2]
# y1<-terra::ext(bc_vect)[3];y2<-terra::ext(bc_vect)[4]
# 
# d_alb_kde<-MASS::kde2d(samp$x, samp$y,n=10000, lims = c(x1,x2,y1,y2))
# #image(d_alb_kde)
# smooth_rast<-rast(d_alb_kde)
# terra::crs(smooth_rast)<-"EPSG:4326"
# ##cropeed
# #croprast<-terra::project(smooth_rast, "EPSG:4326")
# croprast<-terra::crop(smooth_rast, bc_vect)
# croprast<-terra::mask(smooth_rast, bc_vect)
# plot(croprast)
# 
# 
# 
# ggplot() +
#   geom_sf(data = st_transform(bc,4326), color = "lightgrey", alpha = 0.5)+
#   geom_raster(data = as.data.frame(croprast, xy = TRUE),
#               aes(x = x, y = y, fill = z1)) +
#   scale_fill_viridis_c(name = paste0("Presence of ",sppOI))+
#   coord_sf() +
#   #ggthemes::theme_map()+
#   guides(colour = guide_colourbar(position = "right"))+
#   theme(
#     plot.title = element_text(size = rel(2), face = "bold"),
#     plot.subtitle = element_text(size = rel(1.8)),
#     legend.title = element_text(size = rel(1.8)),
#     legend.text = element_text(size = rel(1.2)),
#     legend.position = 'right',
#     panel.background = element_blank(),
#     legend.key.height = unit(1, "null"),
#     axis.text = element_blank(),
#     axis.ticks = element_blank(),
#     axis.title = element_blank()
#   )


```

Goldfish are found in multiple locations, with a concentration in the south west of the province. This could be due to a sampling error, but it more likely due to their method of introduction to British Columbia. Release of goldfish is done by people who had them as pets. It would appear that they are concentrated around areas of high population density.


### Variables {.tabset}

#### Outliers {.tabset}

##### Temperature
```{r temp_plots}

box_temp<-ggplot()+
geom_boxplot(data = samp, aes(y = temp))+
  ylab("Value")+
  labs(title = "Temperature")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )

ggplot()+
geom_boxplot(data = samp_thin, aes(y = temp))+
  labs(title = "Temperature thinned")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )




```

##### Precipitation plots
```{r precip_plot}
box_precip<-ggplot()+
geom_boxplot(data = samp, aes(y = precip))+
  ylab("Value")+
  labs(title = "Precipitation")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )
ggplot()+
geom_boxplot(data = samp_thin, aes(y = precip))+
  labs(title = "Precipitation thinned")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )
```

##### Calcium plots
```{r Ca_plot}
box_ca<-ggplot()+
geom_boxplot(data = samp, aes(y = calc))+
  labs(title = "Calcium")+
  ylab("Value")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )

ggplot()+
geom_boxplot(data = samp_thin, aes(y = calc))+
  ylab("Value")+
  labs(title = "Calcium thinned")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )
```

##### ph plots
```{r pH_plot}
box_ph<-ggplot()+
geom_boxplot(data = samp, aes(y = pH))+
  ylab("Value")+
  labs(title = "pH")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )

ggplot()+
geom_boxplot(data = samp_thin, aes(y = pH))+
  ylab("Value")+
  labs(title = "pH thinned")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )
```

##### Elevation plots
```{r elev_plot}
box_elev<-ggplot()+
geom_boxplot(data = samp, aes(y = elev))+
  ylab("Value")+
  labs(title = "Elev")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank()
        ,
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )

ggplot()+
geom_boxplot(data = samp_thin, aes(y = elev))+
  labs(title = "Elev thinned")+
  theme(plot.title = element_text(size = rel(2), face = "bold"),
        plot.subtitle = element_text(size = rel(1.8)),
        legend.title = element_text(size = rel(1.8)),
        legend.text = element_text(size = rel(1.2)),
        legend.position = 'bottom',
        panel.background = element_blank(),
        axis.text = element_text(size = rel(1.1), face = "bold"),
        axis.title = element_text(face = "bold")
        )
```


```{r}

library(patchwork)

(box_temp | box_precip | box_ca) /
  (box_ph | box_elev)

```


#### Colinearity {.tabset}

##### Pairwise plots - All observations

```{r pairwise_plots}
library(scatterPlotMatrix)
# Z<-cbind(samp$present, samp$temp, samp$precip, samp$calc, samp$pH, samp$elev)
# colnames(Z)<-c("Present", "Temp", "Precip", "Calcium", "pH", "Elevation")
# pairs(Z[,-1], lower.panel = panel.smooth2, upper.panel = panel.cor, diag.panel = panel.hist)

scatterPlotMatrix(samp, zAxisDim = "present", keptColumns = c("temp", "precip", "calc", "pH", "elev"),
                  corrPlotType = "Text")

```


##### Pairwise plots - Thinned

```{r pairwise_plots_thinned}
library(scatterPlotMatrix)
# Z<-cbind(samp$present, samp$temp, samp$precip, samp$calc, samp$pH, samp$elev)
# colnames(Z)<-c("Present", "Temp", "Precip", "Calcium", "pH", "Elevation")
# pairs(Z[,-1], lower.panel = panel.smooth2, upper.panel = panel.cor, diag.panel = panel.hist)

scatterPlotMatrix(samp_thin, zAxisDim = "present", keptColumns = c("temp", "precip", "calc", "pH", "elev"),
                  corrPlotType = "Text")

```

### Correlation of Rasters

```{r Raster_Correlation}
library(ENMTools)

cor<-raster.cor.matrix(stack_data)
thresh<-0.6
dists<-as.dist(1-abs(cor))
clust <- hclust(dists, method = "single")
groups <- cutree(clust, h = 1 - thresh)
## Visualize groups:
plot(clust, hang = -1)
rect.hclust(clust, h = 1 - thresh)
groups
```

The groups displayed here are the names of rasters that are considered to be correlated with each other, subject to a user-defined threshold. In this case, 0.6. Increasing this value will result in smaller groups, which reduces the likelihood of finding correlated variables. Reducing this value further would result in larger groups with more focus on groups that are moderately correlated. If the correlation between the rasters is higher than this, then they are put in the same group. Only one variable should be kept from each group, which would reduce colinearity and increase emphasize biologically important variables. This approach does not deal with multi-colinearity.


### Variance inflation factors {.active}
```{r corvif}
Z<-cbind(samp$present, samp$temp, samp$precip, samp$calc, samp$pH, samp$elev)
colnames(Z)<-c("Present", "Temp", "Precip", "Calcium", "pH", "Elevation")
corvif(Z[, c(-1,-7)])

Z_thin<-cbind(samp_thin$present, samp_thin$temp, samp_thin$precip, samp_thin$calc,
              samp_thin$pH, samp_thin$elev)
colnames(Z_thin)<-c("Present", "Temp", "Precip", "Calcium", "pH", "Elevation")
corvif(Z_thin[, c(-1,-7)])
```

In this, we are looking for variables with values above 3, which is subjective. In this case, temperature, precipiation, pH and elevation are all above this threshold.

### Variance Inflation Factors, 2 ways

```{r}

vif2<-usdm::vif(stack_data)
vif2
```

Documentation for this function suggests values above 10 mean predictors are correlated. The cut-off varies among papers and experts from 5-20 as a suggested cut off point.

<!-- Morans I -->
<!-- This is a measure of spatial autocorrelation. How related are the variables bsedon the locations based on where they are located. -->

```{r morn}

# library(geosphere)
# #samp_dist<-as.matrix(distCosine(c(0,0), cbind(samp$x, samp$y)))
# samp_dist <-distm(samp[1:2], fun = distHaversine)
# samp_dist_inv<-1/samp_dist
# diag(samp_dist_inv)<-0
# samp_dist_inv[is.infinite(samp_dist_inv)]<-0
# 
# 
# Moran.I(samp$temp, samp_dist_inv)
# Moran.I(samp$precip, samp_dist_inv)



```


```{r moran_thin}
# samp_dist <- as.matrix(dist(cbind(samp_thin$x, samp_thin$y)))
# samp_dist<-distm(samp_thin[1:2], fun = distHaversine)
# samp_dist_inv<-1/samp_dist
# diag(samp_dist_inv)<-0
# samp_dist_inv[is.infinite(samp_dist_inv)]<-0
# 
# 
# Moran.I(samp_thin$temp, samp_dist_inv)

```



